They're Clutching up! Team Momentum in Round-Based Esports

### 2022-08-14 paper2/paper2.Rmd
paper 2

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
execute: 
  include: false
  echo: false
  eval: false
format:
  arxiv-pdf:
    keep-tex: true  
  arxiv-html: default
author:
  - name: Tony ElHabr
    affiliations:
      - name: Georgia Institute of Technology
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD).

@@

  Second, I examine post-streak round win probability. I find that teams perform significantly worse than expected after streaks of 2, 3, and 4 wins when series end up going to 9, 10, or 11 (maximum) rounds, even after accounting for the "hot-hand" phenomenon.

  Third, I compare win percentages in round one versus all other rounds, hypothesizing that there may be some advantage on either side when there is no prior information about how the opponent intends to play a given map in either game. I find only one instance for which there seems to be a significant defensive advantage in round one.

  Finally, I evaluate behavior when teams have two rounds left to win the series, observing a peak in COD offensive win percentages in the 4-4 state, and no such oddity in Valorant.
bibliography: references.bib  
---

```{r}
#| label: setup
#| eval: true
library(dplyr)
library(qs)
library(tidyr)
# library(gt)
library(scales)

@@

```{r}
#| label: descriptive-cod-snd-stats
#| eval: true
cod_n_series <- cod_rounds |> 
  distinct(series_id) |> 
  nrow()

@@

Call of Duty (CoD), first released in 2003, is one of the most popular first-person shooter (FPS) video game franchises of all-time. Tthe most popular mode in the competitive scene is "Search and Destroy" (SnD), which bears resemblance to "Bomb Defusal" in Counter-Strike and "Plant/Defuse" in Valorant, two other FPS games played in professional leagues. SnD is one-sided game mode in which one team, the offensive side, tries to destroy one of two designated bomb sites on the map.

In professional CoD, a team must win six rounds of SnD to win the match.[^1] A round can end in one of four ways:

@@

[^2]: Merchanics regarding the bomb:

@@

-   The bomb can be picked up by any member of the offensive team.
-   The bomb carrier is not obstructed at all by carrying the bomb (i.e. movement is the same, weapon usage is the same).
-   The defense does not get any visual indication for who is carrying the bomb.
-   A bomb plant takes five seconds. The timer resets if the player stops planting site prior to completing it.
-   A bomb defuse takes seven seconds. The timer resets if the player "drops" the bomb.
-   The bomb takes 45 seconds to defuse after being planted.

@@

We adopt the terminology "series" to refer to what CoD SnD players typically call a "match", so as to emulate the terminology of playoff series in professional leagues like the National Basketball Association, National Hockey League, and Major League Baseball. A "game" or a "match" in such leagues is analogous to a "round" of CoD SnD.

@@

CoD has roughly gone through three eras of professional gaming: (1) Major League Gaming (MLG) tournaments prior to 2016; (2) the CoD World League (CWL), initiated in 2016; and (3) the 12-franchise CoD League (CDL), running since 2020 and completing three year-long "seasons" completed as of August 2022.[^4] The data set consists of all SnD matches played in tournaments and qualifiers during the CDL era, totaling 7,792 rounds across 852 series. Data was collected in spreadsheets by community member "IOUTurtle".[^5]

@@

[^5]: Data: https://linktr.ee/CDLArchive. Author: https://twitter.com/IOUTurtle

The empirical offensive round win percentage across all rounds is 47.8%.[^6] To contextualize our subsequent discussion of series length and momentum, table 1 shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from this global average.

[^6]: Offensive round win percentage has been nearly constant across the three games during the CDL era: MW (2020) `r pull_cod_o_win_prop_by_game('MW (2020)')`, Cold War (2021) `r pull_cod_o_win_prop_by_game('Cold War (2021)')`, Vanguard (2022) `r pull_cod_o_win_prop_by_game('Vanguard (2022)')`

```{r}
#| label: cod_o_win_prop
#| eval: false
cod_round_and_series_win_prop_by_side <- cod_rounds |>
  group_by(pre_cumu_w, pre_cumu_l, is_offense) |>
  summarize(
    n = n(),
    across(c(win_round, win_series), sum)
  ) |>
  ungroup() |>
  mutate(
    win_round_prop = win_round / n,
    win_series_prop = win_series / n
  )

cod_round_and_series_win_prop_by_side |>
  filter(is_offense) |>
  transmute(
    pre_cumu_w,
    pre_cumu_l,
    lab = sprintf('%.1f%%\n(%s)', round(100 * win_round_prop, 1), n)
  ) |>
  pivot_wider(
    names_from = pre_cumu_l,
    values_from = lab
  ) |>
  arrange(pre_cumu_w) |>
  gt::gt() |>
  gt::cols_label(
    pre_cumu_w = "Defensive team's round wins",
  ) |>
  gt::tab_spanner(
    label = "Offensive team's round wins",
    columns = `0`:`5`
  ) |>
  gt::as_latex()
```

```{=tex}
\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{rrrrrrr}
\toprule
 & \multicolumn{6}{c}{Offensive team's round wins} \\ 
\cmidrule(lr){2-7}
Defensive team's round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
\midrule
0 & 47.8\%
(852) & 46.6\%
(408) & 43.1\%
(216) & 43.5\%
(115) & 43.3\%
(67) & 40.5\%
(37) \\ 
1 & 48.6\%
(444) & 49.3\%
(418) & 51.5\%
(309) & 43.4\%
(205) & 43.3\%
(120) & 39.4\%
(99) \\ 
2 & 52.8\%
(218) & 48.9\%
(305) & 48.9\%
(315) & 46.6\%
(262) & 48.7\%
(189) & 42.1\%
(133) \\ 
3 & 54.5\%
(123) & 46.0\%
(200) & 49.6\%
(250) & 45.6\%
(248) & 44.4\%
(214) & 44.8\%
(174) \\ 
4 & 56.9\%
(65) & 54.5\%
(145) & 47.2\%
(193) & 44.7\%
(228) & 55.2\%
(221) & 50.5\%
(208) \\ 
5 & 47.4\%
(38) & 49.4\%
(83) & 47.1\%
(136) & 50.9\%
(175) & 45.2\%
(177) & 46.0\%
(202) \\ 
\bottomrule
\end{longtable}
```
# Literature review

There have been only a handful of studies of the distribution of games played in a series, most of which assume a constant probability $p$ of a given team winning a game in the series, regardless of the series state. [@mosteller1952] observed that the American League had dominated the National League in Major League Baseball's (MLB) World Series matchups, implying that matchups should not modeled with \$p=0.5\$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $p \approx 0.65$. in each case: (1) a "method of moments" approach in which one solves for $p$. from the empirical average number of games won by the loser of the series; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series, and (3) minimizing the chi-square goodness of fit estimate for $p$.

[@chance2020] re-examines the constant probability notion in Major League Baseball's World Series (1923--2018), the National Basketball Association's Finals (1951--2018), and the National Hockey League's Stanley Cup (1939--2018). Chance applies finds strong evidence against the null hypothesis of $p = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.[^7] Chance's work is closely related to ours and, in fact, provides a guide for the first part of our investigation.

@@

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. [@gilovich1985] do not find evidence for the hot hand phenomenon. Recently , [@miller2018] refute [@gilovich1985]'s conclusions, finding mathematical evidence that seems to support negative recency. Specifically, they find that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" (streak selection bias) that imply that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". We agree with Miller's findings, accounting for the streak selection bias in our study of momentum.

@@

First, we investigate the constant probability assumption and the distribution of rounds played in a series. Afterwards, we investigate momentum, building on our learnings from the constant probability assumption analysis.

@@

[^10]: If $p > 0.5$, then we might say that this team is the better team (presumably known in hindsight)

@@

$$
P_E(i) = \frac{(i - 1)!}{s_1!(i - s_2)!}(p^{s_2}q^{s_2^{'}} + p^{s_2^{'}}q^{s_2}).
$$ {#eq-series-length}

```{r}
#| label: cod_series_outcome_prop
#| eval: true
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_n_games <- function(n, p = 0.5, max_round = .max_round) {
  s <- ceiling(max_round / 2)
  (factorial(n - 1) / (factorial(s - 1) * factorial(n - s))) * (p^s * (1 - p)^(n - s) + p^(n - s) * (1 - p)^s)
}

@@

Table 2 shows the expected proportion for $s = 11$ and the naive assumption that $p = 0.5$, along with the observed proportion $P_{O}(i)$ for CoD SnD.

```{=tex}
\captionsetup[table]{labelformat=empty,skip=1pt}
\begin{longtable}{rrrr}
\toprule
&  & \multicolumn{2}{c}{Observed} \\ 
\cmidrule(lr){3-4}
Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
\midrule
6 & $3.1\%$ & $4.7\%$ & 40 \\ 
7 & $9.4\%$ & $11.9\%$ & 101 \\ 
8 & $16.4\%$ & $16.5\%$ & 141 \\ 
9 & $21.9\%$ & $21.7\%$ & 185 \\ 
10 & $24.6\%$ & $21.5\%$ & 183 \\ 
11 & $24.6\%$ & $23.7\%$ & 202 \\ 
\bottomrule
\end{longtable}
```
The probabilities that a best-of-11 series lasts $i$ rounds ($P(i)$, where $i \in R = [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50% probability ($p = 0.5$) of winning each game {#tbl-prob-series-lasting-i-rounds} along with the observed frequencies of series lasting $i$ rounds, expressed as a percentage $P_O(i)$ and as a count ($N_O$)

```{r}
#| label: cod_series_outcomes_chi
#| eval: false
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

$$
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11]
$$

as `r scales::number(cod_series_outcomes_naive_chi$statistic, accuracy = 0.1)` (p-value of `r scales::number(cod_series_outcomes_naive_chi$p.value, accuracy = 0.0001)`), we can comfortably reject the constant probability null hypothesis, event at a confidence level of $\alpha = 0.01$.

```{r}
#| label: cod-alternative-to-p=0.5-method-prep
#| eval: false
min_p1 <- 0.5
max_p1 <- 0.7
interval_p1 <- 0.0025
cod_ps <- tibble(p = seq(min_p1, max_p1, by = 0.0025))
```

```{r}
#| label: cod-alternative-to-p=0.5-method-1
#| eval: false
# p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
theoretical_cod_series_length <- function(p) {
  6 * prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11) +
    7 * prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11) +
    8 * prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11) +
    9 * prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11) +
    10 * prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11) +
    11 * prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)
}
cod_rounds_per_series <- cod_streaks_naive_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-2
#| eval: false
pluck_cod_n <- function(.n_rounds) {
  cod_streaks_naive_res$series |>
    filter(n_rounds == .n_rounds) |>
    pull(n_actual)
}

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-3
#| eval: false
summarize_cod_series_chi <- function(p) {
  summarize_cod_streaks(p) |> 
    pluck('series') |> 
    perform_series_chi_test()
}

@@

```{r}
#| label: cod-alternatives-to-p=0.5
#| eval: false
cod_streaks_other_res <- tibble(
  p = c(0.5, cod_theoretical_p1, cod_theoretical_p2, cod_theoretical_p3),
  idx = 0L:3L,
  name = c('0. Naive', '1. Method of moments', '2. Maximum likelihood', '3. Minimum Chi-square')
) |> 
  mutate(
    series_streaks = map(
      p,
      ~summarize_cod_streaks(.x) |> 
        pluck('series') 
    ),
    chi_squ = map(
      series_streaks,
      ~.x |> 
        perform_series_chi_test() |> 
        select(statistic, p.value)
    )
  )

@@

@tbl-mosteller-methods shows the alternate values for the constant round win probability $p$ for winning a given round in a CoD SnD that we find when applying the three methods suggested by [@mosteller1952]. Each is approximately or equal to 0.575, and each results in a $chi^2$ value for which we cannot reject constant probability null hypothesis.

| Method                 |    $p$ | $\chi^2$ (p-value) |
|:-----------------------|-------:|-------------------:|
| 0\. Naive              | 0.5000 |     16.0 (\<=0.01) |
| 1\. Method of moments  | 0.5725 |          3.6 (0.6) |
| 2\. Maximum likelihood | 0.5750 |         3.5 (0.62) |
| 3\. Minimum $\chi^2$   | 0.5775 |         3.5 (0.62) |

: Alternate estimates of the constant probability $p$ for winning a given round in a CoD SnD, applying the three methods suggested by [@mosteller1952], in addition to the naive $p = 0.5$. {#tbl-mosteller-methods}

@tbl-expected-series-lengths-alternative-ps shows the new $P_E(i)$ when re-applying @eq-series-length for each new $p$. We observe that $P_E(i)$ is notably smaller for $p_{1,2,3}$ when $i \in [10, 11]$ and higher for $i \in [6, 7]$, more closely matching $P_O(i)$.

| Series lasts $i$ rounds | 0\. $p$= 0.5 | 1\. $p$ = 0.5725 | 2\. $p$ = 0.575 | 3\. $p$ = 0.5775 | $P_O(i)$ |
|------------------------:|-------------:|-----------------:|----------------:|-----------------:|---------:|
|                       6 |         3.1% |             4.1% |            4.2% |             4.3% |     4.7% |
|                       7 |         9.4% |            11.1% |           11.2% |            11.4% |    11.9% |
|                       8 |        16.4% |            17.7% |           17.8% |            17.9% |    16.5% |
|                       9 |        21.9% |            21.8% |           21.8% |            21.8% |    21.7% |
|                      10 |        24.6% |            23.1% |           23.0% |            22.9% |    21.5% |
|                      11 |        24.6% |            22.1% |           22.0% |            21.8% |    23.7% |

: Caption {#tbl-expected-series-lengths-alternative-ps}

>>>>

---
title: "They're Clutching up! Team Momentum in Round-Based Esports"
authors:
  - name: Tony ElHabr
    affiliations:
      - name: Georgia Institute of Technology
    email: anthonyelhabr@gmail.com
    url: https://tonyelhabr.rbind.io
abstract: |
  My research investigates patterns in round win percentages in professional Search and Destroy (SnD) matches of the popular first-person shooter game Call of Duty (CoD).

@@

  Second, I examine post-streak round win probability. I find that teams perform significantly worse than expected after streaks of 2, 3, and 4 wins when series end up going to 9, 10, or 11 (maximum) rounds, even after accounting for the "hot-hand" phenomenon.
bibliography: references.bib  
biblio-style: unsrt
output: rticles::arxiv_article
---

```{r}
#| label: setup
#| include: false
#| echo: false
#| eval: true
knitr::opts_chunk$set(
  eval = FALSE,
  echo = FALSE,
  include = FALSE,
  message = FALSE,
  warning = FALSE
)
```

```{r}
#| label: setup-code
library(dplyr)
library(qs)
library(tidyr)
library(gt)
library(scales)

@@

```{r}
#| label: descriptive-cod-snd-stats
cod_n_series <- cod_rounds |> 
  distinct(series_id) |> 
  nrow()

@@

Call of Duty (CoD), first released in 2003, is one of the most popular first-person shooter (FPS) video game franchises of all-time. The most popular mode in the competitive scene is "Search and Destroy" (SnD), which bears resemblance to "Bomb Defusal" in Counter-Strike and "Plant/Defuse" in Valorant, two other FPS games played in professional leagues. SnD is one-sided game mode in which one team, the offensive side, tries to destroy one of two designated bomb sites on the map.

In professional CoD, a team must win six rounds of SnD to win the match.[^1] A round can end in one of fmy ways:

@@

[^2]:
    -   The bomb can be picked up by any member of the offensive team.
    -   The bomb carrier is not obstructed at all by carrying the bomb (i.e. movement is the same, weapon usage is the same).
    -   The defense does not get any visual indication for who is carrying the bomb.
    -   A bomb plant takes five seconds. The timer resets if the player stops planting site prior to completing it.
    -   A bomb defuse takes seven seconds. The timer resets if the player "drops" the bomb.
    -   The bomb takes 45 seconds to defuse after being planted.

@@

@@

I adopt the terminology "series" to refer to what CoD SnD players typically call a "match", so as to emulate the terminology of playoff series in professional leagues like the National Basketball Association, National Hockey League, and Major League Baseball. A "game" or a "match" in such leagues is analogous to a "round" of CoD SnD.

@@

CoD has roughly gone through three eras of professional gaming: (1) Major League Gaming (MLG) tmynaments prior to 2016; (2) the CoD World League (CWL), initiated in 2016; and (3) the 12-franchise CoD League (CDL), running since 2020 and completing three year-long "seasons" completed as of August 2022.[^4] The data set consists of all SnD matches played in tmynaments and qualifiers during the CDL era, totaling 7,792 rounds across 852 series. Data was collected in spreadsheets by community member "IOUTurtle".[^5]

@@

[^5]: Data: <https://linktr.ee/CDLArchive>. Author: <https://twitter.com/IOUTurtle>

The empirical offensive round win percentage across all rounds is 47.8%.[^6] To contextualize my subsequent discussion of series length and momentum, table \ref{tbl:cod-o-win-prop-by-series-state} shows round win percentages by series "state" (i.e. the number of round wins by each team prior to an upcoming round). Offensive round win rate is not quite constant, although never veers more than 10% from this global average.

[^6]: Offensive round win percentage has been nearly constant across the three games during the CDL era: 1. 47.2% in MW (2020) 2. 47.9% in Cold War (2021) 3. 48.1% in Vanguard (2022)

```{r}
#| label: cod_o_win_prop
cod_round_and_series_win_prop_by_side <- cod_rounds |>
  group_by(pre_cumu_w, pre_cumu_l, is_offense) |>
  summarize(
    n = n(),
    across(c(win_round, win_series), sum)
  ) |>
  ungroup() |>
  mutate(
    win_round_prop = win_round / n,
    win_series_prop = win_series / n
  )

cod_round_and_series_win_prop_by_side |>
  filter(is_offense) |>
  transmute(
    pre_cumu_w,
    pre_cumu_l,
    lab = sprintf('%.1f%%\n(%s)', round(100 * win_round_prop, 1), n)
  ) |>
  pivot_wider(
    names_from = pre_cumu_l,
    values_from = lab
  ) |>
  arrange(pre_cumu_w) |>
  gt::gt() |>
  gt::cols_label(
    pre_cumu_w = "Defense round wins",
  ) |>
  gt::tab_spanner(
    label = "Offense round wins",
    columns = `0`:`5`
  ) |>
  gt::as_latex()
```

```{=tex}
\begin{longtable}{crrrrrr}
  \caption{Offensive round win rates for the upcoming round, given both the offensive and defensive team's prior number of round wins}\label{tbl:cod-o-win-prop-by-series-state} \\
  \toprule
   & \multicolumn{6}{c}{Offense round wins} \\ 
  \cmidrule(lr){2-7}
  Defense round wins & 0 & 1 & 2 & 3 & 4 & 5 \\ 
    \midrule
    0 & 47.8\%
    (852) & 46.6\%
    (408) & 43.1\%
    (216) & 43.5\%
    (115) & 43.3\%
    (67) & 40.5\%
    (37) \\ 
    1 & 48.6\%
    (444) & 49.3\%
    (418) & 51.5\%
    (309) & 43.4\%
    (205) & 43.3\%
    (120) & 39.4\%
    (99) \\ 
    2 & 52.8\%
    (218) & 48.9\%
    (305) & 48.9\%
    (315) & 46.6\%
    (262) & 48.7\%
    (189) & 42.1\%
    (133) \\ 
    3 & 54.5\%
    (123) & 46.0\%
    (200) & 49.6\%
    (250) & 45.6\%
    (248) & 44.4\%
    (214) & 44.8\%
    (174) \\ 
    4 & 56.9\%
    (65) & 54.5\%
    (145) & 47.2\%
    (193) & 44.7\%
    (228) & 55.2\%
    (221) & 50.5\%
    (208) \\ 
    5 & 47.4\%
    (38) & 49.4\%
    (83) & 47.1\%
    (136) & 50.9\%
    (175) & 45.2\%
    (177) & 46.0\%
    (202) \\ 
  \bottomrule
\end{longtable}
```
# Literature review

There have been only a handful of studies of the distribution of games played in a series, most of which assume a constant probability $p$ of a given team winning a game in the series, regardless of the series state. [@mosteller1952] observed that the American League had dominated the National League in Major League Baseball's (MLB) World Series matchups, implying that matchups should not modeled with $p = 0.5$. Mosteller proposed three approaches for identifying the optimal constant probability value of the stronger team in the World Series, finding $p \approx 0.65$. in each case: (1) a "method of moments" approach in which one solves for $p$. from the empirical average number of games won by the loser of the series; (2) maximizing the likelihood that the sample would have been drawn from a population in which the probability of a team winning a game is constant across the series, and (3) minimizing the chi-square goodness of fit estimate for $p$.

[@chance2020] re-examines the constant probability notion in Major League Baseball's World Series (1923--2018), the National Basketball Association's Finals (1951--2018), and the National Hockey League's Stanley Cup (1939--2018). Chance applies finds strong evidence against the null hypothesis of $p = 0.5$ in the MLB and NHL championship series when applying Mosteller's first and second methods.[^7] Chance's work is closely related to mys and, in fact, provides a guide for the first part of my investigation.

@@

Studying both player streaks and team streaks in basketball, in both observational and controlled settings. [@gilovich1985] do not find evidence for the hot hand phenomenon. Recently , [@miller2018] refute [@gilovich1985]'s conclusions, finding mathematical evidence that seems to support negative recency. Specifically, they find that a "bias exists in a common measure of the conditional dependence of present outcomes on streaks of past outcomes in sequential data" (streak selection bias) that imply that, under i.i.d. conditions, "the proportion of successes among the outcomes that immediately follow a streak of consecutive successes is expected to be strictly less than the underlying (conditional) probability of success". I agree with Miller's findings, accounting for the streak selection bias in my study of momentum.

@@

First, I investigate the constant probability assumption and the distribution of rounds played in a series. Afterwards, I investigate momentum, building on my learnings from the constant probability assumption analysis.

@@

[^10]: If $p > 0.5$, then I might say that this team is the better team (known omnisciently).

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq-series-length}{}{
P_E(i) = \frac{(i - 1)!}{s_1!(i - s_2)!}(p^{s_2}q^{s_2^{'}} + p^{s_2^{'}}q^{s_2}).
}\label{eq:series-length}\end{equation}
```
```{r}
#| label: cod_series_outcome_prop
.max_round <- 11
.cutoff <- 6
prob_of_series_lasting_n_games <- function(n, p = 0.5, max_round = .max_round) {
  s <- ceiling(max_round / 2)
  (factorial(n - 1) / (factorial(s - 1) * factorial(n - s))) * (p^s * (1 - p)^(n - s) + p^(n - s) * (1 - p)^s)
}

@@

Table \ref{tbl:cod-prob-series-lasting-i-rounds} shows the expected proportion for $s = 11$ and the naive assumption that $p = 0.5$, along with the observed frequencies observed in CoD SnD, as both a count $N_O(i)$ (of series lasting $i$ rounds) and a proportion $P_O(i)$ of all series.

```{=tex}
\begin{longtable}{rrrr}
  \caption{The probabilities that a best-of-11 series lasts $i$ rounds ($Pr(i)$, where $i = {6, 7, 8, 9, 10, 11}$) under the assumption that each team has a 50% probability ($p = 0.5$) of winning each round.}\label{tbl:cod-prob-series-lasting-i-rounds} \\
  \toprule
  &  & \multicolumn{2}{c}{Observed} \\ 
  \cmidrule(lr){3-4}
  Series lasts $i$ rounds & $P_E(i)$ & $P_O(i)$ & $N_O(i)$ \\ 
  \midrule
  6 & $3.1\%$ & $4.7\%$ & 40 \\ 
  7 & $9.4\%$ & $11.9\%$ & 101 \\ 
  8 & $16.4\%$ & $16.5\%$ & 141 \\ 
  9 & $21.9\%$ & $21.7\%$ & 185 \\ 
  10 & $24.6\%$ & $21.5\%$ & 183 \\ 
  11 & $24.6\%$ & $23.7\%$ & 202 \\ 
  \bottomrule
\end{longtable}
```
The probabilities that a best-of-11 series lasts $i$ rounds ($P(i)$, where $i \in R = [6, 7, 8, 9, 10, 11]$) under the assumption that each team has a 50% probability ($p = 0.5$) of winning each game {#tbl-prob-series-lasting-i-rounds} along with the observed frequencies of series lasting $i$ rounds, expressed as a percentage $P_O(i)$ and as a count ($N_O$)

```{r}
#| label: cod_series_outcomes_naive_chi
generate_chi_label <- function(statistic, p.value) {
  sprintf('%.1f (%s)', statistic, ifelse(p.value <= 0.01, '<=0.01', as.character(round(p.value, 2))))
}

@@

```{=tex}
\begin{equation}\protect\hypertarget{eq-series-length}{}{
\chi^2 = \sum^{11}_{i=6} \frac{(P_O(i) - P_E(i))^2}{P_E(i)}, i \in R = [6, 7, 8, 9, 10, 11].
}\label{eq:chi-squ-test}\end{equation}
```
as 16.0 (p-value of 0.0068), I can comfortably reject the constant probability null hypothesis, event at a confidence level of $\alpha = 0.01$.

```{r}
#| label: cod-alternative-to-p=0.5-method-prep
min_p1 <- 0.5
max_p1 <- 0.7
interval_p1 <- 0.0025
cod_ps <- tibble(p = seq(min_p1, max_p1, by = 0.0025))
```

```{r}
#| label: cod-alternative-to-p=0.5-method-1
## p. 365 on https://math.mit.edu/classes/18.095/2016IAP/lec9/Sports_Mosteller1952_WorldSeries.pdf
theoretical_cod_series_length <- function(p) {
  6 * prob_of_series_lasting_n_games(n = 6, p = p, max_round = 11) +
    7 * prob_of_series_lasting_n_games(n = 7, p = p, max_round = 11) +
    8 * prob_of_series_lasting_n_games(n = 8, p = p, max_round = 11) +
    9 * prob_of_series_lasting_n_games(n = 9, p = p, max_round = 11) +
    10 * prob_of_series_lasting_n_games(n = 10, p = p, max_round = 11) +
    11 * prob_of_series_lasting_n_games(n = 11, p = p, max_round = 11)
}
cod_rounds_per_series <- cod_streaks_naive_res$series |>
  summarize(
    actual = sum(n_rounds * prop_actual),
    expected = sum(n_rounds * prop_expected)
  )

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-2
pluck_cod_n <- function(.n_rounds) {
  cod_streaks_naive_res$series |>
    filter(n_rounds == .n_rounds) |>
    pull(n_actual)
}

@@

```{r}
#| label: cod-alternative-to-p=0.5-method-3
summarize_cod_series_chi <- function(p) {
  summarize_cod_streaks(p) |> 
    pluck('series') |> 
    perform_series_chi_test()
}

@@

```{r}
#| label: cod-alternatives-to-p=0.5
cod_streaks_other_res <- tibble(
  p = c(0.5, cod_theoretical_p1, cod_theoretical_p2, cod_theoretical_p3),
  idx = 0L:3L,
  name = c('0. Naive', '1. Method of moments', '2. Maximum likelihood', '3. Minimum Chi-square')
) |> 
  mutate(
    series_streaks = map(
      p,
      ~summarize_cod_streaks(.x) |> 
        pluck('series') 
    ),
    chi_squ = map(
      series_streaks,
      ~.x |> 
        perform_series_chi_test() |> 
        select(statistic, p.value)
    )
  )

@@

Table \ref{tbl:mosteller-methods-results} shows the alternate values for the constant round win probability $p$ for winning a given round in a CoD SnD that I find when applying the three methods suggested by Mosteller [-@mosteller1952]. Each is approximately or equal to 0.575, and, when applying Equation \ref{eq:chi-squ-test}, each results in a $\chi^2$ value for which I cannot reject the constant probability null hypothesis.

```{=tex}
\begin{longtable}[]{@{}lrr@{}}
  \caption{Alternate estimates of the constant probability \(p\) for winning a given round in a CoD SnD, applying the three methods suggested by Mosteller (1952), in addition to the naive \(p = 0.5\).}\label{tbl:mosteller-methods-results} \\
  \toprule()
    Method & \(p\) & \(\chi^2\) (p-value) \\
    \midrule()
    \endfirsthead
    \toprule()
    Method & \(p\) & \(\chi^2\) (p-value) \\
    \midrule()
  \endhead
  0. Naive & 0.5000 & 16.0 (\textless=0.01) \\
  1. Method of moments & 0.5725 & 3.6 (0.6) \\
  2. Maximum likelihood & 0.5750 & 3.5 (0.62) \\
  3. Minimum \(\chi^2\) & 0.5775 & 3.5 (0.62) \\
  \bottomrule()
\end{longtable}
```
Table \ref{tbl:expected-series-lengths-alternative-ps} shows the new $P_E(i)$ when re-applying Equation \ref{eq:series-length} for each new $p$. I observe that $P_E(i)$ is notably smaller for $p_{1,2,3}$ when $i \in [10, 11]$ and higher for $i \in [6, 7]$, more closely matching $P_O(i)$.

```{=tex}
\begin{longtable}[]{@{}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.2427}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1456}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1748}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1650}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.1748}}
  >{\raggedleft\arraybackslash}p{(\columnwidth - 10\tabcolsep) * \real{0.0971}}@{}}
\caption{The probabilities that a best-of-11 series lasts $i$ rounds under various assumptions for $p$.}\label{tbl:expected-series-lengths-alternative-ps} \\
\toprule()
\begin{minipage}[b]{\linewidth}\raggedleft
Series lasts \(i\) rounds
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
0. \(p\) = 0.5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
1. \(p\) = 0.5725
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
2. \(p\) = 0.575
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
3. \(p\) = 0.5775
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(P_O(i)\)
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedleft
Series lasts \(i\) rounds
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
0. \(p\) = 0.5
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
1. \(p\) = 0.5725
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
2. \(p\) = 0.575
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
3. \(p\) = 0.5775
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedleft
\(P_O(i)\)
\end{minipage} \\
\midrule()
\endhead
6 & 3.1\% & 4.1\% & 4.2\% & 4.3\% & 4.7\% \\
7 & 9.4\% & 11.1\% & 11.2\% & 11.4\% & 11.9\% \\
8 & 16.4\% & 17.7\% & 17.8\% & 17.9\% & 16.5\% \\
9 & 21.9\% & 21.8\% & 21.8\% & 21.8\% & 21.7\% \\
10 & 24.6\% & 23.1\% & 23.0\% & 22.9\% & 21.5\% \\
11 & 24.6\% & 22.1\% & 22.0\% & 21.8\% & 23.7\% \\
\bottomrule()
\end{longtable}
```
So I conclude that the constant round win probability assumption can be valid in CoD SnD series with the appropriate choice of $p$ ($\approx 0.575$).
